\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-08-26 paper.qmd
Updated Figure 16 caption

The MNIST dataset consists of $70000$ grayscale images of handwritten digits [@lecun2010]. @yingfan2021 used this dataset to evaluate how PaCMAP preserves local structures in \gD{}. Because of the large number of images, the focus was specifically on the handwritten digit 1, which exhibits a non-linear structure in \gD{} (@fig-pacmap-author). There are $7877$ images of the digit 1 in the dataset. Additionally, PCA was applied as a preprocessing step and the first $10$ principal components were selected. The objective was to assess whether PaCMAP effectively preserves the non-linear structure in \gD{}.

@@

<!--PaCMAP param: n_components=2, n_neighbors=10, init=random, MN_ratio=0.9, FP_ratio=2.0-->
```{r}
#| fig-cap: "$2\\text{-}D$ layout from PaCMAP applied for the digit 1 of the MNIST dataset. The (hyper-)parameter settings, beyond the defaults, are $10$ nearest neighbors, ratio of the number of mid-near pairs to the number of neighbors equal to $0.5$, and the ratio of the number of further pairs to the number of neighbors is $2$. We use our *model-in-the-data-space* to assess whether this is an accurate representation of structure present in $10\\text{-}D$, or if it is misleading. The angle of the digit 1 images varies along this structure. Images at the top-left of the $2\\text{-}D$ layout show the digit 1 angled more to the left, while those at the bottom-right show the digit 1 angled more to the right."
#| label: fig-pacmap-author
#| fig-pos: H
#| out-height: 22%

@@

<!--bin1 = 22, bin2 = 17, b = 374, non_empty = 140-->
```{r}
#| echo: false
#| fig-cap: "$10\\text{-}D$ model error in $2\\text{-}D$. The $2\\text{-}D$ model shows a non-linear continuous structure. Most low $10\\text{-}D$ model errors are distributed along the lower edge of the $2\\text{-}D$ structure, while most large $10\\text{-}D$ model errors are concentrated along the upper edge."
#| label: fig-model-mnist
#| fig-pos: H
#| out-height: 22%

According to @fig-mnist1-sc1, the non-linear continuous structure observed in the \gD{} representation of PaCMAP (@fig-pacmap-author) is also visible when visualizing the model overlaid on the data space. This indicates that PaCMAP accurately captures the structure of the $10\text{-}D$ data. Additionally, the model shows a twisted pattern within the non-linear structure in $10\text{-}D$ space (@fig-mnist1-sc2), which is an additional pattern not visible in the \gD{} representation (@fig-pacmap-author). Furthermore, as shown in @fig-mnist1-sc3, some long edges exist in the $10\text{-}D$ space that are not recognized as long edges in the \gD{} representation. However, PaCMAP is a reasonable \gD{} representation of MNIST digit 1 data, because PaCMAP preserves the non-linear structure present in the $10\text{-}D$ data.

@@

<!--add langevitour screenshots and youtube animation link-->
```{r}
#| label: fig-mnist-tri-proj
#| fig-pos: H
#| fig-height: 10
#| fig-width: 10
#| fig-cap: "Model in \\gD{}, on the PaCMAP layout, and three views of the fit in projections from $10\\text{-}D$, for the digit 1 of MNIST data ($(s_1, \ s_2) = (-0.100, \ -0.059)$, $b = 374 \  (22, \ 17)$, $m = 140$, benchmark value to remove large edges is $0.094$). (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/zcg_GXBmqjA>)."

@@

There are certain data points that exhibit high error rates due to their deviation from the usual $10\text{-}D$ data structure, which makes them anomalies (@fig-model-mnist b). These anomalies can be classified into two types: those that are anomalies within the non-linear structure and those that lie outside of it. The images associated with high model error points within the non-linear structure display different patterns of the digit 1, as shown in @fig-mnist1-within. However, when comparing these images to the ones found outside of the non-linear structure, it becomes evident that the latter display different patterns of the digit 1 (@fig-mnist1-out).

>>>>

The MNIST dataset consists of $70000$ grayscale images of handwritten digits [@lecun2010]. @yingfan2021 used this dataset to evaluate how PaCMAP preserves local structures in \gD{}. Because of the large number of images, the focus was specifically on the handwritten digit 1, which exhibits a non-linear structure in \gD{} (@fig-pacmap-author). There are $7877$ images of the digit 1 in the dataset. Additionally, PCA was applied as a preprocessing step and the first $10$ principal components were selected. The objective is to assess whether PaCMAP effectively preserves the non-linear structure in \gD{}.

@@

<!--PaCMAP param: n_components=2, n_neighbors=10, init=random, MN_ratio=0.9, FP_ratio=2.0-->
```{r}
#| fig-cap: "$2\\text{-}D$ layout from PaCMAP applied for the digit 1 of the MNIST dataset. The (hyper-)parameter settings, beyond the defaults, are $10$ nearest neighbors, ratio of the number of mid-near pairs to the number of neighbors equal to $0.5$, and the ratio of the number of further pairs to the number of neighbors is $2$. We use our *model-in-the-data-space* to assess whether this is an accurate representation of non-linear structure present in $10\\text{-}D$, or if it is misleading. The angle of the digit 1 images varies along this structure. Images at the top-left of the $2\\text{-}D$ layout show the digit 1 angled more to the left, while those at the bottom-right show the digit 1 angled more to the right."
#| label: fig-pacmap-author
#| fig-pos: H
#| out-height: 22%

@@

<!--bin1 = 22, bin2 = 17, b = 374, non_empty = 140-->
According to @fig-mnist-tri-proj (b), the non-linear structure observed in the \gD{} layout of PaCMAP (@fig-pacmap-author) is also visible when visualizing the model overlaid on the data space. This indicates that PaCMAP accurately captures the non-linear structure of the $10\text{-}D$ data. Additionally, the model shows a twisted pattern within the non-linear structure in $10\text{-}D$ space (@fig-mnist-tri-proj (c)), which is an additional pattern not visible in the \gD{} representation (@fig-pacmap-author). Furthermore, as shown in @fig-mnist-tri-proj (d), some long edges exist in the $10\text{-}D$ space that are not recognized as long edges in the \gD{} representation. However, PaCMAP provides a reasonable \gD{} representation of MNIST digit 1 data, as it preserves the non-linear structure present in the $10\text{-}D$ data.

@@

<!--add langevitour screenshots and youtube animation link-->
```{r}
#| label: fig-mnist-tri-proj
#| fig-pos: H
#| fig-height: 10
#| fig-width: 10
#| fig-cap: "Model in \\gD{}, on the PaCMAP layout, and three views of the fit in projections from $10\\text{-}D$, for the digit 1 of MNIST data ($a_1 = 0.062$, $b = 285 \\  (19, \ 15)$, $m = 109$, and benchmark value to remove large edges is $0.108$). (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/zcg_GXBmqjA>)."

@@

```{r}
#| echo: false
#| fig-cap: "$10\\text{-}D$ model error in $2\\text{-}D$. The $2\\text{-}D$ model shows a non-linear continuous structure. Most low $10\\text{-}D$ model errors are distributed along the lower edge of the $2\\text{-}D$ structure, while most large $10\\text{-}D$ model errors are concentrated along the upper edge."
#| label: fig-model-mnist
#| fig-pos: H
#| out-height: 22%

There are certain data points that exhibit high error rates due to their deviation from the usual $10\text{-}D$ data structure, which makes them anomalies (@fig-model-mnist). These anomalies can be classified into two types: those that are anomalies within the non-linear structure and those that lie outside of it. The images associated with high model error points within the non-linear structure display different patterns of the digit 1, as shown in @fig-mnist1-within. However, when comparing these images to the ones found outside of the non-linear structure, it becomes evident that the latter display different patterns of the digit 1 (@fig-mnist1-out).

### 2024-08-26 paper.tex
Updated Figure 16 caption

The MNIST dataset consists of \(70000\) grayscale images of handwritten
digits \citep{lecun2010}. \citet{yingfan2021} used this dataset to
evaluate how PaCMAP preserves local structures in \gD{}. Because of the
large number of images, the focus was specifically on the handwritten
digit 1, which exhibits a non-linear structure in \gD{}
(Figure~\ref{fig-pacmap-author}). There are \(7877\) images of the digit
1 in the dataset. Additionally, PCA was applied as a preprocessing step
and the first \(10\) principal components were selected. The objective
was to assess whether PaCMAP effectively preserves the non-linear
structure in \gD{}.

@@

\caption{\label{fig-pacmap-author}\(2\text{-}D\) layout from PaCMAP
applied for the digit 1 of the MNIST dataset. The (hyper-)parameter
settings, beyond the defaults, are \(10\) nearest neighbors, ratio of
the number of mid-near pairs to the number of neighbors equal to
\(0.5\), and the ratio of the number of further pairs to the number of
neighbors is \(2\). We use our \emph{model-in-the-data-space} to assess
whether this is an accurate representation of structure present in
\(10\text{-}D\), or if it is misleading. The angle of the digit 1 images
varies along this structure. Images at the top-left of the
\(2\text{-}D\) layout show the digit 1 angled more to the left, while
those at the bottom-right show the digit 1 angled more to the right.}

@@

According to \textbf{?@fig-mnist1-sc1}, the non-linear continuous
structure observed in the \gD{} representation of PaCMAP
(Figure~\ref{fig-pacmap-author}) is also visible when visualizing the
model overlaid on the data space. This indicates that PaCMAP accurately
captures the structure of the \(10\text{-}D\) data. Additionally, the
model shows a twisted pattern within the non-linear structure in
\(10\text{-}D\) space (\textbf{?@fig-mnist1-sc2}), which is an
additional pattern not visible in the \gD{} representation
(Figure~\ref{fig-pacmap-author}). Furthermore, as shown in
\textbf{?@fig-mnist1-sc3}, some long edges exist in the \(10\text{-}D\)
space that are not recognized as long edges in the \gD{} representation.
However, PaCMAP is a reasonable \gD{} representation of MNIST digit 1
data, because PaCMAP preserves the non-linear structure present in the
\(10\text{-}D\) data.

\caption{\label{fig-mnist-tri-proj}Model in \gD{}, on the PaCMAP layout,
and three views of the fit in projections from \(10\text{-}D\), for the
digit 1 of MNIST data (\((s_1,  s_2) = (-0.100,  -0.059)\),
\(b = 374   (22,  17)\), \(m = 140\), benchmark value to remove large
edges is \(0.094\)). (The \textbf{langevitour} software is used to view
the data with a tour, and the full video is available at
\url{https://youtu.be/zcg_GXBmqjA}).}

There are certain data points that exhibit high error rates due to their
deviation from the usual \(10\text{-}D\) data structure, which makes
them anomalies (Figure~\ref{fig-model-mnist} b). These anomalies can be
classified into two types: those that are anomalies within the
non-linear structure and those that lie outside of it. The images
associated with high model error points within the non-linear structure
display different patterns of the digit 1, as shown in
Figure~\ref{fig-mnist1-within}. However, when comparing these images to
the ones found outside of the non-linear structure, it becomes evident
that the latter display different patterns of the digit 1
(Figure~\ref{fig-mnist1-out}).

>>>>

The MNIST dataset consists of \(70000\) grayscale images of handwritten
digits \citep{lecun2010}. \citet{yingfan2021} used this dataset to
evaluate how PaCMAP preserves local structures in \gD{}. Because of the
large number of images, the focus was specifically on the handwritten
digit 1, which exhibits a non-linear structure in \gD{}
(Figure~\ref{fig-pacmap-author}). There are \(7877\) images of the digit
1 in the dataset. Additionally, PCA was applied as a preprocessing step
and the first \(10\) principal components were selected. The objective
is to assess whether PaCMAP effectively preserves the non-linear
structure in \gD{}.

@@

\caption{\label{fig-pacmap-author}\(2\text{-}D\) layout from PaCMAP
applied for the digit 1 of the MNIST dataset. The (hyper-)parameter
settings, beyond the defaults, are \(10\) nearest neighbors, ratio of
the number of mid-near pairs to the number of neighbors equal to
\(0.5\), and the ratio of the number of further pairs to the number of
neighbors is \(2\). We use our \emph{model-in-the-data-space} to assess
whether this is an accurate representation of non-linear structure
present in \(10\text{-}D\), or if it is misleading. The angle of the
digit 1 images varies along this structure. Images at the top-left of
the \(2\text{-}D\) layout show the digit 1 angled more to the left,
while those at the bottom-right show the digit 1 angled more to the
right.}

According to Figure~\ref{fig-mnist-tri-proj} (b), the non-linear
structure observed in the \gD{} layout of PaCMAP
(Figure~\ref{fig-pacmap-author}) is also visible when visualizing the
model overlaid on the data space. This indicates that PaCMAP accurately
captures the non-linear structure of the \(10\text{-}D\) data.
Additionally, the model shows a twisted pattern within the non-linear
structure in \(10\text{-}D\) space (Figure~\ref{fig-mnist-tri-proj}
(c)), which is an additional pattern not visible in the \gD{}
representation (Figure~\ref{fig-pacmap-author}). Furthermore, as shown
in Figure~\ref{fig-mnist-tri-proj} (d), some long edges exist in the
\(10\text{-}D\) space that are not recognized as long edges in the \gD{}
representation. However, PaCMAP provides a reasonable \gD{}
representation of MNIST digit 1 data, as it preserves the non-linear
structure present in the \(10\text{-}D\) data.

\caption{\label{fig-mnist-tri-proj}Model in \gD{}, on the PaCMAP layout,
and three views of the fit in projections from \(10\text{-}D\), for the
digit 1 of MNIST data (\(a_1 = 0.062\), \(b = 285 \  (19,  15)\),
\(m = 109\), and benchmark value to remove large edges is \(0.108\)).
(The \textbf{langevitour} software is used to view the data with a tour,
and the full video is available at \url{https://youtu.be/zcg_GXBmqjA}).}

@@

There are certain data points that exhibit high error rates due to their
deviation from the usual \(10\text{-}D\) data structure, which makes
them anomalies (Figure~\ref{fig-model-mnist}). These anomalies can be
classified into two types: those that are anomalies within the
non-linear structure and those that lie outside of it. The images
associated with high model error points within the non-linear structure
display different patterns of the digit 1, as shown in
Figure~\ref{fig-mnist1-within}. However, when comparing these images to
the ones found outside of the non-linear structure, it becomes evident
that the latter display different patterns of the digit 1
(Figure~\ref{fig-mnist1-out}).
